Higher-Order unification for free!

### 2024-04-29 paper/main.tex
minor

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial and does not only concern the head of rules, but the entire sequence
of unification problems that constitute the execution of a logic program.
In fact
the solution for \elpiIn{P} above generates a
(Coq) $\beta$-redex in the second premise (the predicate
under the \elpiIn{pi w\ }\hspace{-0.4em}).
% We show below the premise before and
% after the instantiation of \elpiIn{P}:

% \begin{elpicode}
% decision (app [                   P                  , w])
% decision (app [lam A (a\ app [con"nfact", a, con"3"]), w])
% \end{elpicode}

% \noindent
In turn, this redex prevents rule \ref{clause:r2} from backchaining properly since
the following unification problem has no solution:

@@

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program in \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the higher-order unification procedure of the meta language \Ue{} can be
efficiently used to simulate a higher-order unification procedure \Uo for
the object language that features $\eta\beta$-conversion. We show how \Uo
can be extended with heuristics to deal with problems outside the pattern
fragment.\\
Section~\ref{sec:problem-statement} states the problem formally and gives the
intuition behind our solution; \cref{sec:simulation} sets up a basic
simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
extend it to higher-order logic programs in the pattern fragment
while \cref{sec:beta} goes beyond the pattern fragment.
Section \ref{sec:implementation} discusses the implementation in Elpi.\\
The $\lambda$Prolog code discussed in the paper can be accessed at the
address \githubUrl.

>>>>

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial and does not only concern the head of rules, but the entire sequence
of unification problems that constitute the execution of a logic program.

In fact
the solution for \elpiIn{P} above generates a
(Coq) $\beta$-redex in the second premise (the predicate
under the \elpiIn{pi w\ }\hspace{-0.4em}).
% We show below the premise before and
% after the instantiation of \elpiIn{P}:
% \begin{elpicode}
% decision (app [                   P                  , w])
% decision (app [lam A (a\ app [con"nfact", a, con"3"]), w])
% \end{elpicode}
% \noindent
In turn, this redex prevents rule \ref{clause:r2} from backchaining properly since
the following unification problem has no solution:

@@

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program in \Fo{} to a strongly related
logic program in \Ho, the language of the meta language, and we show that
the higher-order unification procedure of the meta language \Ue{} can be
used to simulate a higher-order unification procedure \Uo for
the object language that features $\eta\beta$-conversion. We show how \Uo
can be extended with heuristics to deal with problems outside the pattern
fragment.\\
Section~\ref{sec:problem-statement} states the problem formally and gives the
intuition behind our solution; \cref{sec:simulation} sets up a basic
simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
extend it to higher-order logic programs in the pattern fragment
while \cref{sec:beta} goes beyond the pattern fragment.
Section \ref{sec:implementation} discusses the implementation in Elpi.\\
The $\lambda$Prolog code discussed in the paper can be accessed at the
address \githubUrl.
